---
title: Jump-starting visualization with dittoViz
author:
- name: Daniel Bunis
  affiliation: Data Science CoLab, University of California, San Francisco
  email: daniel.bunis@ucsf.edu
date: "Jan 3rd, 2024"
output:
  md_document:
    toc: true
bibliography: ref.bib
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE,
    fig.path = "images/",
    dev="png", dpi = 150, fig.width = 4.5, fig.height = 3.5)
```

# Introduction

dittoViz is a tool built to make analysis and visualization of data easier,
faster, and more accessible. All included visualizations were built with
color-vision impaired users and viewers in mind. Thus, it provides functions for
jump-starting many useful visualization types, which all utilize red-green
color-blindness optimized colors by default, and which allow sufficient
customization, via discrete inputs, for out-of-the-box creation of
publication-ready figures.

The intention is for dittoViz to be a generalized version of the popular yet
scRNAseq focused dittoSeq package. Although we have started with simply pulling
the plotting features out of all dittoSeq's visualizations, the goal is to
include many more plot types and to grow dittoViz into a comprehensive,
data-type agnostic visualization suite. If there is a plot type missing that
you would like to suggest be added, please do so by
[creating an issue](https://github.com/dtm2451/dittoViz/issues/new) on the
github!

## Color-blindness friendliness:

The default colors of this package are red-green color-blindness friendly. To
make it so, I used the suggested colors from [@wong_points_2011] and adapted
them slightly by appending darker and lighter versions to create a 24 color
vector. All plotting functions use these colors, stored in `dittoColors()`, by
default.

Additionally:

- Shapes displayed in the legends are generally enlarged as this can be almost
as helpful as the actual color choice for colorblind individuals.
- When sensible, dittoViz functions have a shape.by input for having groups
displayed through shapes in addition to color. (But note: even as a red-green
color impaired individual myself writing this vignette, I often still recommend
using color and generally use shapes as either a duplication of the color-data
or for showing additional groupings.)
- scatterPlots can be generated with letters overlaid (set do.letter = TRUE)

# Installation

dittoViz will eventually be available through CRAN.  This section will be
updated once that is the case.

For now, install from the GitHub:

```{r, eval=FALSE}
# Install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

# Install dittoViz
remotes::install_github("dtm2451/dittoViz")
```

# Getting started

## Input Data

All dittoViz functions have a `data_frame` input which takes a data.frame-like
object with observations / individual data points in rows, and features of those
observations in columns.

To get started here, we'll make use of the `?dittoExampleData` documentation
point which was created entirely in order to allow single-line, fast, generation
<!-- of an example `data_frame` input: -->

```{r}
library(dittoViz)
example("dittoExampleData", echo = FALSE)

head(example_df, 10)
summary(example_df)
```

# Visualizations

There are many different types of dittoViz visualizations. Each has intuitive
defaults which allow creation of immediately usable plots. Each also has many
additional tweaks available through discrete inputs that can help ensure you
can create precisely-tuned, deliberately-labeled, publication-quality plots
directly 'out-of-the-box'!

## scatterPlot

These show data points overlaid on a scatter plot, using two numeric columns of
the `data_frame` as the axes and optionally additional columns for color, shape,
faceting, or other features.

```{r, results = "hold"}
# Simplest Form
scatterPlot(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2"
)
```

```{r}
# Additionally colored by a discrete or numeric feature 
scatterPlot(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "clustering")
scatterPlot(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "gene1")
```

```{r}
# Additionally shaped or faceted by a discrete feature 
scatterPlot(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "clustering",
    shape.by = "clustering")
scatterPlot(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "clustering",
    split.by = "conditions")
```

### Additional features

Various additional features can be overlaid on top of these plots.
Adding each is controlled by an input that starts with `add.` or `do.` such as:

- `do.label`
- `do.ellipse`
- `do.letter`
- `do.contour`
- `do.hover`
- `add.trajectory.by.groups`
- `add.trajectory.curves`

Additional inputs that apply to and adjust these features will then start with
the XXXX part that comes after `add.XXXX` or `do.XXXX`, as exemplified below.
(Tab-completion friendly!)

A few examples:

```{r}
# A bit noisy looking because we're adding a LOT of the extras here...
scatterPlot(
    example_df, x.by = "PC1", y.by = "PC2", color.by = "clustering",
    
    do.label = TRUE,
    labels.repel = FALSE,
    
    do.ellipse = TRUE,
    
    do.contour = TRUE,
    contour.color = "gray90",
    
    add.trajectory.by.groups = list(
        c(1,2),
        c(1,4,3)),
    trajectory.group.by = "clustering")
```

## yPlot (+ ridgePlot, ridgeJitter, boxPlot wrappers)

This plotter displays **continuous** data on a continuous-axis,
grouped in the other axis by any discrete feature such as sample or condition.

Data can be represented with violin plots, box plots, jitter representing
individual data points, and/or ridge plots, and the `plots` input controls which
of these data representations are used.
(The function name can be a bit of a misnomer in that ridge plots will use the
x-axis for the continuous data, and the y-axis for your groupings.)

The `group.by` input controls how the data are grouped in the x-axis.

The optional `color.by` input controls the colors that fill in violin, box, and
ridge plots.

```{r, results = "hold"}
yPlot(
    example_df, "gene1", group.by = "clustering"
)
```

`yPlot()` is the main function, but `ridgePlot()`, `ridgeJitter()`, and
`boxPlot()` are wrappers which just adjust the default for the `plots` input
from c("vlnplot", "boxplot", "jitter") to c("ridgeplot"),
c("ridgeplot", "jitter"), or c("boxplot","jitter"), respectively.

```{r}
ridgePlot(example_df, "gene1", group.by = "clustering")
ridgeJitter(example_df, "gene1", group.by = "clustering")
boxPlot(example_df, "gene1", group.by = "clustering")
```

### Adjustments to data representations

Tweaks to the individual data representation types can be made with discrete
inputs, all of which start with the representation types' name.  For
example...

```{r}
yPlot(example_df, "gene1", group.by = "clustering",
    plots = c("vlnplot", "jitter", "boxplot"), # <- order matters
    
    # change the color and size of jitter points
    jitter.color = "blue", jitter.size = 0.5, jitter.width = 1,
    
    # change the outline color and width, and remove the fill of boxplots
    boxplot.color = "white", boxplot.width = 0.1,
    boxplot.fill = FALSE,
    
    # change how the violin plot widths are normalized across groups
    vlnplot.scaling = "count",
    # also add lines at specified quantiles
    vlnplot.quantiles = c(0.1, 0.9),
    
    # add a dotted line representing some important cutoff
    add.line = 0.9
    )
```

## barPlot & freqPlot

These functions quantify and display frequencies of values of a **discrete**
data per sample / clustering / some other discrete data.

For both, data can be represented as either percentages or counts, and this is
controlled by the `scale` input.

```{r, results = "hold"}
# barPlot
barPlot(example_df, var = "sample", group.by = "clustering")
barPlot(example_df, var = "sample", group.by = "clustering",
    scale = "count")
```

freqPlot separates each 'var'-value into its own facet, and thus puts more
emphasis on each individual element. An additional `sample.by` input controls
splitting of cells within `group.by`-groups into individual samples.

```{r, results = "hold"}
# freqPlot
freqPlot(
    data_frame = example_df,
    var = "clustering",
    sample.by = "sample",
    group.by = "category")
```

## scatterHex

An alternative to scatterPlot where instead of plotting each data point
individually, observations are first binned into hexagonal regions, and then a
summary from all observations of each bin is plotted.

The plot type is most handy for very densely packed data, and allows plotting 1)
observation density alone via color, or 2) observation density via opacity in
addition to some other feature via color.

```{r, results = "hold"}
# Simplest Form, Density by color
scatterHex(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2"
)
```

```{r}
# Colored instead via a discrete or numeric feature 
scatterHex(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "clustering")
scatterHex(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "gene1")
```

### Summary methods

One key point of control lies how the `color.by` data are summarized. This is
controlled with the `color.method` input. As we saw above, this input defaults
to the "max" or most common value for discrete `color.by`-data, and to the
median value for numeric `color.by` data.

For discrete `color.by` data, we also offer `color.method = "max.prop"` for
highlighting diversity among regions of the plot.

For continuous `color.by` data, any named function which inputs a vector of
numbers and outputs a single numberic value can be used.  E.g. mean, min, max,
sum, or sd (standard deviation) might all be useful in a given context.

```{r}
scatterHex(
    data_frame = example_df,
    x.by = "PC1", y.by = "PC2",
    color.by = "gene1",
    color.method = "sum")
```

### Additional features

All additional features of scatterPlot, that don't rely on plotting of all
individual data points, are available here as well:

- `do.label`
- `do.ellipse`
- `do.contour`
- `do.hover`
- `add.trajectory.by.groups`
- `add.trajectory.curves`

A few examples:

```{r}
# A bit noisy looking because we're adding a LOT of the extras here...
scatterHex(
    example_df, x.by = "PC1", y.by = "PC2", color.by = "clustering",
    
    do.label = TRUE,
    labels.repel = FALSE,
    
    do.ellipse = TRUE,
    
    do.contour = TRUE,
    contour.color = "gray90",
    
    add.trajectory.by.groups = list(
        c(1,2),
        c(1,4,3)),
    trajectory.group.by = "clustering")
```

# Customization via Simple Inputs 

**Many adjustments can be made with simple additional inputs**.  Here, we'll go
through a few that are consistent across most dittoViz functions, but there
are many more.  Be sure to check the function documentation (e.g.
`?scatterPlot`) to explore more!  Often, there will be a section towards the
bottom of a function's documentation dedicated to its specific tweaks!

## Subsetting to certain observations / rows

The data shown in a given plot can be adjusted with the `rows.use`
input. This can be provided as either a list of rownames to
include, as an integer vector with the indices of which rows to keep, or as a
logical vector that states whether each row should be included.

```{r}
# Original
barPlot(example_df, var = "sample", group.by = "clustering", scale = "count")

# First 10 cells
barPlot(example_df, var = "sample", group.by = "clustering", scale = "count",
    # String method
    rows.use = c("obs1", "obs2", "obs3", "obs4", "obs5")
    # Index method, which would achieve the same effect
    # rows.use = 1:5
    )

# "1"-cluster only 
barPlot(example_df, var = "sample", group.by = "clustering", scale = "count",
    # Logical method
    rows.use = example_df$clustering == 1)
```

## Faceting with split.by

Most dittoViz plot types can be faceted into separate plots based on the values
of 1 or 2 discrete columns with a `split.by` input.

```{r}
yPlot(example_df, "gene1", group.by = "clustering",
    split.by = "category",
    sub = "faceted by category")
scatterPlot(example_df, x.by = "PC1", y.by = "PC2", color.by = "gene1", 
    split.by = c("category", "subcategory"),
    sub = "faceted by category and subcategory")
```

Extra control over how this is done can be achieved with the `split.adjust`
input. `split.adjust` allows inputs to be passed through to the ggplot 
functions used for achieving the faceting.

```{r}
yPlot(example_df, "gene1", group.by = "clustering", split.by = "category",
    split.adjust = list(scales = "free_y"))
```

When splitting is by only one metadata, the shape of the facet grid can be
controlled with `split.ncol` and `split.nrow`.

```{r, fig.height=7}
ridgePlot(example_df, "gene1", group.by = "clustering", split.by = "category",
    split.ncol = 1)
```

## All titles are adjustable.

Relevant inputs are generally `main`, `sub`, `xlab`, `ylab`, and `legend.title`.

```{r}
barPlot(example_df, "sample", group.by = "clustering",
    main = "Encounters",
    sub = "By Type",
    xlab = NULL, # NULL = remove
    ylab = "Generation 1",
    legend.title = "Types",
    x.labels.rotate = FALSE)
```

## Colors can be adjusted easily.

Colors are normally set with `color.panel` or `max.color` and `min.color`.
When color.panel is used (discrete data), an additional input called `colors`
sets the order in which those are actually used to make swapping around colors
easy when nearby groups appear too similar in scatter plots!

```{r, results="hold"}
# original - discrete
scatterPlot(example_df, "PC1", "PC2", "clustering")
# swapped colors
scatterPlot(example_df, "PC1", "PC2", "clustering",
    colors = 4:1)
# different colors
scatterPlot(example_df, "PC1", "PC2", "clustering",
    color.panel = c("red", "orange", "purple", "yellow", "skyblue"))
```

```{r, results="hold"}
# original - continuous
scatterPlot(example_df, "PC1", "PC2", "gene1")
# different colors
scatterPlot(example_df, "PC1", "PC2", "gene1",
    max.color = "red", min.color = "gray90")
```

## Underlying data can be output.

Simply add  `data.out = TRUE` to any of the individual plotters and a
representation of the underlying data will be output.

```{r}
barPlot(example_df, "sample", group.by = "clustering",
    data.out = TRUE)
```

## plotly hovering can be added.

Many dittoViz functions can be supplied `do.hover = TRUE` to have them convert
the output into an interactive plotly object that will display additional data
about each data point when the user hovers their cursor on top.

Generally, a second input, `hover.data`, is used for denoting what extra
data to display. This input takes in a vector of column names in the order you
wish for them to be displayed. However, when the types of underlying data
possible to be shown are constrained because the plot pieces represent summary
data (barPlot and freqPlot), the `hover.data` input is not used.

```{r, eval = FALSE}
# These can be finicky to render in knitting, but still, example code:
scatterPlot(example_df, "PC1", "PC2", "gene1",
    do.hover = TRUE,
    hover.data = c("gene1", "sample", "clustering", "timepoint"))
barPlot(example_df, "sample", group.by = "clustering",
    do.hover = TRUE)
```

## Rasterization / flattening to pixels

Sometimes, datasets have so many data points that working with each one rendered
individually becomes prohibitively computationally intensive, especially when
opening them in a vector-based graphics editor, such as Illustrator. In such
instances, it can be helpful to have the per-point graphics layers flattened
to a pixel representation. Generally, dittoViz offers this capability via
`do.raster` and `raster.dpi` inputs.

```{r}
# Note: dpi gets re-set by the styling code of this vignette, so this is
#   just a code example -- the rendered plot won't look quite as intended.
scatterPlot(example_df, "PC1", "PC2", "gene1",
    do.raster = TRUE,
    raster.dpi = 300)
```

# The 'colLevels()' Helper Function

dittoViz includes one helper functions that make it easier to determine the
discrete values within a column of your data frame. There are base R functions
which can achieve the same purpose, but the utility of this function is it works
regardless of whether the target data is a character, numeric, or factor type.

This functionality is useful for certain inputs, ex: `vars.use` of `freqPlot()`,
for determining the full option set which you might want to pull only a few
options from.

```{r}
# Tells the unique values of any discrete data column
# The 'conditions' column is a factor, while 'clustering' is a character vector
colLevels("conditions", example_df)
colLevels("clustering", example_df)
```

There are 2 addition inputs for the function, `rows.use` & `used.only`:

You can use the function along with standard dittoViz `rows.use` subsetting to
see if any groupings are removed by the given subsetting.

```{r}
# We can use rows.use to achieve any subsetting prior to value/level summary.
colLevels(
    "conditions", example_df,
    rows.use = example_df$conditions!='condition1'
)
```

And if you wish to know the recorded levels of your data, regardless of whether
any data points exist for those levels, you can set `used.only` to FALSE.
`used.only` is TRUE by default.

```{r}
# Note: Set 'used.only' (default = TRUE) to FALSE to show unused levels
#  of data that are already factors.  By default, only the used options
#  of the data are given.
colLevels("conditions", example_df,
    rows.use = example_df$conditions!="condition1"
)
colLevels("conditions", example_df,
    rows.use = example_df$conditions!="condition1",
    used.only = FALSE
)
```

# Session information

```{r}
sessionInfo()
```

# References
